Countering language attrition with PanLex and the Web of Data

### 2013-01-02 paper/2012/SWJ-PanLex/paper.tex

\section{Linking}
\label{sec:linking}
The Sparqlify view in the previous section ~\ref{fig:ex:sparqlify-ml} already
established the interlinking of the PanLex languages with Lexvo.
In this section we outline the interlinking with DBpedia.

@@

%Since Lexvo covers all ISO 639-3 language codes, interlinking could be done with
%the Sparqlify-ML view in Figure~\ref{fig:ex:sparqlify-ml}.
For DBpedia we were interested in creating \emph{valid} and
thus \emph{dereferenceable} links.
Therefore, we iterated the \emph{titles} datasets\footnote{\url{http://wiki.dbpedia.org/Downloads38} - Note that more languages are available in the raw download section at \url{http://downloads.dbpedia.org/3.8/}},
which map (non-localized) DBpedia URIs to their page titles in the respective language.
For each language version we normalized the labels by applying
Unicode NFKD\footnote{\url{http://unicode.org/reports/tr15/}} normalization and
removal of punctuation characters.
% PanLex now uses NFC, not NFKD, for character normalization and does not remove punctuation.
Each DBpedia resource was then mapped to the PanLex expression that was equal to the resource's normalized label in the respective language.
Table ~\ref{fig:plx-dbp-link-counts} summarizes the number of links obtained.
\begin{table}
\centering
\begin{scriptsize}
\begin{tabular}{lllr}
\toprule
Language & 639-1 & 639-3 & Links \\
\midrule
English    & en & eng & 1,415,241 \\
German     & de & deu &   224,146 \\
French     & fr & fra &   187,364 \\
Italian    & it & ita &   147,485 \\
Spanish    & sp & spa &   117,056 \\
Portuguese & pt & por &   112,266 \\
Polish     & pl & pol &   110,974 \\
Russian    & ru & rus &    68,040 \\
Czech      & cs & ces &    28,767 \\
Catalan    & ca & cat &    27,779 \\
Korean     & ko & kor &    24,912 \\
Turkish    & tr & tur &    22,258 \\
Bulgarian  & bg & bul &    19,431 \\
Hungarian  & hu & hun &    18,203 \\
Slovene    & sl & slv &    11,981 \\
Greek      & el & ell &     1,112 \\
\midrule
Total      &    &     & 2,537,015 \\
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Number of DBpedia links per language}
\label{fig:plx-dbp-link-counts}
\end{table}

@@

With the RDF version of PanLex, a large lexical resource participates in the Web of Data,
making it available to a broader community. Users can easily explore the PanLex data using
the services provided by SNORQL and Pubby. This also increases the chance of development of
new interesting mashups.
For example, the TeraDict translation service\footnote{\url{http://panlex.org/cgi-bin/apidemo-2.cgi}}
could now be easily realized using simple SPARQL queries.
% as shown in
%\todo{Create the query: translations in a given target language of an expression}. 
Another future usage scenario is to link PanLex to Wortschatz with the aim of 
improving the recall of word-by-word translations by considering the co-occurences.

\section{Related Work}
\label{sec:related}
PanLex is an integration project of many existing lexical resources with
the goal of eventually covering all languages in the world, especially
ones having few speakers.
The extraction of information from linguistic sources, and
techniques for automatically inferring translations, are
relevant work discussed in \cite{panlex_probtrans}.

An important initiative is the
\emph{Global Wordnet Association}\footnote{\url{http://www.globalwordnet.org/}},
which offers a platform for sharing wordnets and defines several goals.
These include setting forth standards for uniformly representing wordnets
of different languages and establishing a universal index of meaning.

%It draws from the EuroWordNet~\cite{eurowordnet} and the  
%Manually engineered dictionaries %such as EuroWordNet [38]
% andcurrently mainly used for research purposes\todo{correct?}.
%A list of the resources used in panlex is compiled on the website\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.
%In this section we discuss work related to the extraction and inference
%approaches used in PanLex, and work on the Semantic Web.

%- There has been considerable research on methods to acquire translation lexicons from either
%MRDs (Neff and McCord, 1990; Helmreich et
%al., 1993; Copestake et al., 1994) or from par-
%allel text (Gale and Church, 1991; Fung, 1995;
%Melamed, 1997; Franz et al., 2001), but this has
%generally been limited to a small number of lan-
%guages.
%- Manually engineered dictionaries such as
%EuroWordNet (Vossen, 1998) (whats the name of the new project?)
%- TRANS GRAPH algorithm (Etzioni etal., 2007).

At the level of the Semantic Web, there is currently a trend in publishing linguistic resources
as Linked Data\footnote{\url{http://www.w3.org/DesignIssues/LinkedData.html}} and via SPARQL{\footnote{\url{http://www.w3.org/TR/sparql11-query/}}}.
These efforts are referred to as the \emph{Linguistic Linked Open Data Cloud} (LLOD)\footnote{\url{http://linguistics.okfn.org/resources/llod/}}. 
%Some of these datasets already make use of quasi-standard linguistic Semantic Web Ontolgies, such as  
Several (quasi-)standard ontologies have been developed for covering different aspects of linguistic resources.
Examples include the \emph{Ontologies of Linguistic Annotation} (OLiA)~\cite{olia2010},
the \emph{Lexicon Model for Ontologies} (lemon)~\cite{lemon2011} for modeling lexicon and machine-readable dictionaries,
\emph{POWLA} for modeling linguistic corpora\cite{powla2012}
and the \emph{Natural Language Processing Interchange Format} (NIF)\footnote{\url{http://nlp2rdf.org/nif-1-0}}.

>>>>

\section{Linking}
\label{sec:linking}
The Sparqlify view in the previous section (\autoref{fig:ex:sparqlify-ml}) already
established the interlinking of the PanLex languages with Lexvo.
In this section we outline the interlinking with DBpedia.

@@

%Since Lexvo covers all ISO 639-3 language codes, interlinking could be done with
%the Sparqlify-ML view in Figure~\ref{fig:ex:sparqlify-ml}.
For DBpedia we were interested in creating \emph{valid} and
thus \emph{dereferenceable} links.
Therefore, we iterated the \emph{titles} datasets\footnote{\url{http://wiki.dbpedia.org/Downloads38} - Note that more languages are available in the raw download section at \url{http://downloads.dbpedia.org/3.8/}},
which map (non-localized) DBpedia URIs to their page titles in the respective language.
For each language version we normalized the labels by applying
Unicode NFKD\footnote{\url{http://unicode.org/reports/tr15/}} normalization and
removal of punctuation characters.
% PanLex now uses NFC, not NFKD, for character normalization and does not remove punctuation.
Each DBpedia resource was then mapped to the PanLex expression that was equal to the resource's normalized label in the respective language.
\autoref{fig:plx-dbp-link-counts} summarizes the number of links obtained.
\begin{table}
\centering
\begin{scriptsize}
\begin{tabular}{lllr}
\toprule
Language & 639-1 & 639-3 & Links \\
\midrule
English    & en & eng & 1,415,241 \\
German     & de & deu &   224,146 \\
French     & fr & fra &   187,364 \\
Italian    & it & ita &   147,485 \\
Spanish    & sp & spa &   117,056 \\
Portuguese & pt & por &   112,266 \\
Polish     & pl & pol &   110,974 \\
Russian    & ru & rus &    68,040 \\
Czech      & cs & ces &    28,767 \\
Catalan    & ca & cat &    27,779 \\
Korean     & ko & kor &    24,912 \\
Turkish    & tr & tur &    22,258 \\
Bulgarian  & bg & bul &    19,431 \\
Hungarian  & hu & hun &    18,203 \\
Slovene    & sl & slv &    11,981 \\
Greek      & el & ell &     1,112 \\
\midrule
Total      &    &     & 2,537,015 \\
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Number of DBpedia links per language}
\label{fig:plx-dbp-link-counts}
\end{table}

@@

With the RDF version of PanLex, a large lexical resource participates in the Web of Data,
making it available to a broader community. Users can easily explore the PanLex data using
the services provided by SNORQL\footnote{\url{https://github.com/kurtjx/SNORQL}} and Pubby\footnote{\url{http://wifo5-03.informatik.uni-mannheim.de/pubby/}}. This also increases the chance of development of
new interesting mashups.
For example, the TeraDict translation service\footnote{\url{http://panlex.org/cgi-bin/apidemo-2.cgi}}
could now be easily realized using simple SPARQL queries.
% as shown in
%\todo{Create the query: translations in a given target language of an expression}. 
Another future usage scenario is to link PanLex to Wortschatz\footnote{\url{http://wortschatz.uni-leipzig.de/}} with the aim of 
improving the recall of word-by-word translations by considering the co-occurences.

\section{Related Work}
\label{sec:related}
PanLex is an integration project of many existing lexical resources. % with
% the goal of eventually covering all languages in the world, especially
% ones having few speakers.
The extraction of information from linguistic sources, and
techniques for automatically inferring translations, are
relevant work discussed in \cite{panlex_probtrans}.

An important initiative is the
\emph{Global Wordnet Association}\footnote{\url{http://www.globalwordnet.org/}},
which offers a platform for sharing wordnets and defines several goals.
These include setting forth standards for uniformly representing wordnets
of different languages and establishing a universal index of meaning.
%
%It draws from the EuroWordNet~\cite{eurowordnet} and the  
%Manually engineered dictionaries %such as EuroWordNet [38]
% andcurrently mainly used for research purposes\todo{correct?}.
%A list of the resources used in panlex is compiled on the website\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.
%In this section we discuss work related to the extraction and inference
%approaches used in PanLex, and work on the Semantic Web. 
%
%- There has been considerable research on methods to acquire translation lexicons from either
%MRDs (Neff and McCord, 1990; Helmreich et
%al., 1993; Copestake et al., 1994) or from par-
%allel text (Gale and Church, 1991; Fung, 1995;
%Melamed, 1997; Franz et al., 2001), but this has
%generally been limited to a small number of lan-
%guages.
%- Manually engineered dictionaries such as
%EuroWordNet (Vossen, 1998) (whats the name of the new project?)
%- TRANS GRAPH algorithm (Etzioni etal., 2007).
%
At the level of the Semantic Web, there is currently a trend in publishing linguistic resources
%as Linked Data\footnote{\url{http://www.w3.org/DesignIssues/LinkedData.html}} and via SPARQL{\footnote{\url{http://www.w3.org/TR/sparql11-query/}}}.
as Linked Data\footnote{\url{http://www.w3.org/DesignIssues/LinkedData.html}} and via SPARQL.
These efforts are referred to as the \emph{Linguistic Linked Open Data Cloud} (LLOD). 
%Some of these datasets already make use of quasi-standard linguistic Semantic Web Ontolgies, such as  
Several (quasi-)standard ontologies have been developed for covering different aspects of linguistic resources.
Examples include the \emph{Ontologies of Linguistic Annotation} (OLiA)~\cite{olia2010},
the \emph{Lexicon Model for Ontologies} (lemon)~\cite{lemon2011} for modeling lexicon and machine-readable dictionaries,
\emph{POWLA} for modeling linguistic corpora\cite{powla2012}
and the \emph{Natural Language Processing Interchange Format} (NIF)\footnote{\url{http://nlp2rdf.org/nif-1-0}}.
